---
output: html_document
---

```{r message=FALSE}
library(tidyverse)
library(vegan)
library(ape)
```

I'm providing a few customized functions to avoid some tricky business with 
reading a distance matrix and running a PCoA. I know, it's cheating.

```{r}
tibble_to_distmat <- function (x) {
  x <- as.data.frame(x)
  rownames(x) <- x[,1]
  x <- x[,-1]
  as.dist(x)
}
custom_pcoa <- function (d) {
  pc <- pcoa(d)
  pcdf <- pc$vectors %>%
    as.data.frame() %>%
    as_tibble(rownames = "sample_id") %>%
    select(sample_id, Axis.1, Axis.2, Axis.3, Axis.4)
  pcdf
}
```

This is my extreme shorthand for variable names:

  * `s` is always the table of samples
  * `cts` is always the table of read counts
  * `uu` is always the unweighted UniFrac distance matrix
  * `wu` is always the weighted UniFrac distance matrix

```{r message=FALSE}
s <- read_csv("mouse_vendors_samples.csv")
cts <- read_csv("mouse_vendors_otu_table.csv")
uu <- read_csv("mouse_vendors_unweighted_unifrac.csv") %>%
  tibble_to_distmat()
wu <- read_csv("mouse_vendors_weighted_unifrac.csv") %>%
  tibble_to_distmat()
```

# Number of samples

```{r}
s %>%
  count(sample_type, vendor)
```

```{r}
s %>%
  count(subject_id, sample_type) %>%
  ggplot() +
  geom_tile(aes(x=sample_type, y=subject_id, fill=n))
```

# Number of reads

```{r}
cts_total <- cts %>%
  group_by(sample_id) %>%
  summarize(total_reads = sum(counts)) %>%
  ungroup()
```

```{r}
cts_total %>%
  ggplot() +
  geom_histogram(aes(x=total_reads), binwidth = 1000, boundary = 0)
```

# Alpha diversity

```{r}
cts_alpha <- cts %>%
  group_by(sample_id) %>%
  summarize(richness = rarefy(counts, 5000), shannon = diversity(counts))
```

```{r}
s_alpha <- s %>%
  left_join(cts_alpha, by="sample_id")
```

```{r}
s_alpha %>%
  ggplot() +
  geom_boxplot(aes(x=vendor, y=richness, fill=sample_type))
```

# Beta diversity

```{r}
uu_pcoa <- custom_pcoa(uu)
uu_pcoa %>%
  left_join(s, by="sample_id") %>%
  ggplot() +
  geom_point(aes(x=Axis.1, y=Axis.2, shape=sample_type, color=vendor))
```

```{r}
set.seed(42)
adonis(uu ~ vendor + sample_type, data=s)
```

# Differential abundance

```{r}
props <- cts %>%
  group_by(sample_id) %>%
  mutate(proportion = counts / sum(counts)) %>%
  ungroup()
```

```{r}
taxon_props <- props %>%
  group_by(sample_id, taxon) %>%
  summarize(counts = sum(counts), proportion = sum(proportion)) %>%
  ungroup()
```

```{r}
taxon_props %>%
  group_by(taxon) %>%
  filter(mean(proportion) > 0.01) %>%
  ungroup() %>%
  ggplot() +
  geom_tile(aes(x=sample_id, y=taxon, fill = proportion)) +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1))
```


```{r}
top_taxa <- taxon_props %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(proportion)) %>%
  ungroup() %>%
  arrange(desc(mean_prop)) %>%
  slice(1:10)
```

```{r}
taxon_props %>%
  filter(taxon %in% "Bacteroidetes - S24-7") %>%
  left_join(s, by="sample_id") %>%
  # Add a very small number to proportions to prevent zeros
  mutate(proportion = proportion + 1e-6) %>%
  ggplot() +
  geom_boxplot(aes(x=vendor, y=proportion, fill=sample_type)) +
  scale_y_log10()
```

```{r}
taxon_props %>%
  filter(taxon %in% "Bacteroidetes - S24-7") %>%
  left_join(s, by="sample_id") %>%
  filter(sample_type %in% "Stool") %>%
  mutate(log_prop = log10(proportion + 1e-6)) %>%
  lm(log_prop ~ vendor, data=.) %>%
  summary()
```

